Confined Cubic Blue Phases under Shear

### 2011-11-15 papers/lmcproc/lmcproc.tex
Removed 2 todos
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1489 40876441-3496-0410-a613-a2fca35b2ae2

% TODO
% OLIVER: Please check Olmsted reference: Mike says ``*not* in this paper''
% at point of reference
% DAVIDE (?): Page 6, sentence ``It is interesting to note that the
% secondary flow in the BP does not match the direction...'' Mike says
% ``incomprehensible''
% ANYONE: Please add reference ''to explain strange p_0/rho bit in
% equation 5''. 
% OLIVER (lucky you): Please adjust to use standard rheological x,y,z
% Cartesian coordinates. (y,z,x) at moment?

@@

The time evolution of the fluid density and velocity are respectively governed
by the continuity equation
$\partial_t \rho = -\partial_\alpha(\rho u_\alpha)$, and
the following Navier-Stokes equation:
\begin{eqnarray}
\partial_t u_\alpha +\rho \,u_\beta \partial_\beta u_\alpha
= \partial_\beta \Pi_{\alpha \beta}
+\eta\, \partial_\beta \{ \partial_\alpha u_\beta + \partial_\beta u_\alpha
+(1+3\frac{P_0}{\rho} )\partial_\mu u_\mu \delta_{\alpha \beta}\}. 
\label{eqn6}
\end{eqnarray}
At low flow rates the fluid can be considered as incompressible, so that the
last term on the right hand side of Eq.~\ref{eqn6} remains small.
$\eta$ is an isotropic background viscosity which is set to $\eta=0.8333$ in our case.
The pressure tensor reads explicitly
\begin{eqnarray}
\Pi_{\alpha \beta}&=&P_0-\xi H_{\alpha \gamma}\left(Q_{\gamma \beta} +\frac{1}{3} \delta_{\gamma \beta}\right)-\xi \left(Q_{\alpha \gamma} +\frac{1}{3} \delta_{\alpha \gamma}\right) H_{\gamma \beta}\nonumber\\
&+& 2 \xi  \left(Q_{\alpha \beta} +\frac{1}{3} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}-\partial_\alpha Q_{\gamma \nu} \frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
&+&Q_{\alpha \gamma}H_{\gamma \beta}-H_{\alpha \gamma} Q_{\gamma \beta}.
\label{eqn7}
\end{eqnarray}
In the isotropic state ${\bf Q}\equiv 0$ and Eq.\ref{eqn7} is reduced to the
scalar pressure which, in a system at rest, is constant to a very good
approximation.

@@

It is interesting to compare the results for different gap sizes.
For $n_x=1$ we observe that the BPII-network is slightly deformed with
respect to the quiescent state, but always reforms at the same position.
For $n_x > 1$, the BPII network distorts more under the same shear rate, 
suggesting that the structure within the sample is strongly affected by
the anchoring at the boundaries.
Interestingly, when the sample is thicker than one unit cell, a movement
of the network in the negative vorticity direction arises
which is not observed for $n_x=1$.
This network motion does not proceed at a constant speed, but rather
proceeds in a ``stick-slip'' fashion: elastic stresses first slowly build up,
and are then quickly released as the network ``slips'' along the vorticity
direction. Changing from a left-handed to a right-handed helix inverts the
direction of motion of the network.
Hence the ``stick-slip'' motion is related to the chiral nature of the helix.
This phenomenon may bear some similarity to the travelling wave motion in
cholesterics sheared along the vorticity direction 
below a critical Ericksen number \cite{Rey:1996a, Rey:1996b}.
It is interesting to note that the secondary flow in the BP
does {\em not} match the direction of the ``stick-slip'' motion of the network, which is then not connected to mass flow
(in this respect this phenomenon is close to a permeation flow).

@@

Our simulations show that the disclination network of BPII continuously breaks and reforms while the network undergoes an affine deformation under shear. This process leads to an oscillatory shear stress in time, although the oscillations are only regular for gap sizes of a single unit cell across the flow gradient direction, where anchoring affects the physics of the system greatly, in agreement with recent simulations of quiescent BP networks \cite{Fukuda:2010a, Fukuda:2010b, Ravnik:2011b}.
For larger gap sizes of two and more unit cells a ``stick-slip'' motion of
the network arises, along the vorticity direction. The associated network
migration is not directly linked to mass flow, and is therefore somewhat
akin to permeation flow in cholesterics. The average speed of this network
migration depends on the thermodynamic parameters, and appears to be higher
for higher temperature and lower chirality.

>>>>

% TODO
% OLIVER: Please check Olmsted reference: Mike says ``*not* in this paper''
% at point of reference
% OLIVER (lucky you): Please adjust to use standard rheological x,y,z
% Cartesian coordinates. (y,z,x) at moment?

@@

The time evolution of the fluid density and velocity are respectively governed
by the continuity equation
$\partial_t \rho = -\partial_\alpha(\rho u_\alpha)$, and
the following Navier-Stokes equation:
\begin{eqnarray}
\partial_t u_\alpha +\rho \,u_\beta \partial_\beta u_\alpha
= \partial_\beta \Pi_{\alpha \beta}
+\eta\, \partial_\beta \{ \partial_\alpha u_\beta + \partial_\beta u_\alpha
+(1+3\frac{\partial P_0}{\partial\rho} )\partial_\mu u_\mu \delta_{\alpha \beta}\}. 
\label{eqn6}
\end{eqnarray}
The final term in Eq.~\ref{eqn6} arises from the Chapman-Enskog expansion
of the LB equations \cite{Denniston:2001}.
At low flow rates the fluid can be considered as incompressible, so that the
last term on the right hand side of Eq.~\ref{eqn6} remains small.
$\eta$ is an isotropic background viscosity which is set to $\eta=0.8333$ in our case.
The pressure tensor reads explicitly
\begin{eqnarray}
\Pi_{\alpha \beta}&=&P_0 \delta_{\alpha\beta}
-\xi H_{\alpha \gamma}\left(Q_{\gamma \beta} +\frac{1}{3} \delta_{\gamma \beta}\right)-\xi \left(Q_{\alpha \gamma} +\frac{1}{3} \delta_{\alpha \gamma}\right) H_{\gamma \beta}\nonumber\\
&+& 2 \xi  \left(Q_{\alpha \beta} +\frac{1}{3} \delta_{\alpha \beta}\right) Q_{\gamma \nu} H_{\gamma \nu}-\partial_\alpha Q_{\gamma \nu} \frac{\delta{\cal F}}{\delta \partial_{\beta} Q_{\gamma \nu}}\nonumber\\
&+&Q_{\alpha \gamma}H_{\gamma \beta}-H_{\alpha \gamma} Q_{\gamma \beta}.
\label{eqn7}
\end{eqnarray}
In the isotropic state ${\bf Q}\equiv 0$ and Eq.\ref{eqn7} is reduced to the
scalar pressure which, in a system at rest, is constant to a very good
approximation.

@@

It is interesting to compare the results for different gap sizes.
For $n_x=1$ we observe that the BPII-network is slightly deformed with
respect to the quiescent state, but always reforms at the same position.
For $n_x > 1$, the BPII network distorts more under the same shear rate, 
suggesting that the structure within the sample is strongly affected by
the anchoring at the boundaries.
Interestingly, when the sample is thicker than one unit cell, a movement
of the network in the negative vorticity direction arises
which is not observed for $n_x=1$.
This network motion does not proceed at a constant speed, but rather
proceeds in a ``stick-slip'' fashion: elastic stresses first slowly build up,
and are then quickly released as the network ``slips'' along the vorticity
direction. Changing from a left-handed to a right-handed helix inverts the
direction of motion of the network.
Hence the ``stick-slip'' motion is related to the chiral nature of the helix.
This phenomenon may bear some similarity to the travelling wave motion in
cholesterics sheared along the vorticity direction 
below a critical Ericksen number \cite{Rey:1996a, Rey:1996b}.

@@

Our simulations show that the disclination network of BPII continuously breaks and reforms while the network undergoes an affine deformation under shear. This process leads to an oscillatory shear stress in time, although the oscillations are only regular for gap sizes of a single unit cell across the flow gradient direction, where anchoring affects the physics of the system greatly, in agreement with recent simulations of quiescent BP networks \cite{Fukuda:2010a, Fukuda:2010b, Ravnik:2011b}.
For larger gap sizes (two and more unit cells) a ``stick-slip'' motion of
the network arises along the vorticity direction. The associated network
migration is not directly linked to mass flow, and is therefore somewhat
akin to permeation flow in cholesterics. The average speed of this network
migration depends on the thermodynamic parameters, and appears to be higher
for higher temperature and lower chirality.
